Storm surges in the Strait of Georgia simulated with a regional model

### 2015-05-14 AO_guides/manuscriptAO.tex
addressed all referee comments

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modular tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. {\color{red} Although these features are not being used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia.} This work focuses on the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study. {\color{red} NEMO has been successfully used in other coastal modelling studies in the Northeast Atlantic \citep{maraldi2013nemo} and Mediterranean (CITATION)!}

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait (Figure \ref{fig:domain}). Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells {\color{red} inward from the open boundary}. As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis. {\color{red} A magnified image of the complex passages through the San Jaun and Gulf Islands is provided in Figure \ref{fig:zoom}.}

@@

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. {\color{red}Tidal predictions were generated with a tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge. In order to prevent over-predicted constituents due to high surge years, the time series is first filtered by applying a Doodson filter \citep{parker2007NOAA} and removing periods with large non-tidal amplitudes. Long period constituents such as SA, SSA, etc are not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected. } The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

{\color{red} Although Tofino is not the closest tide guage to the western boundary, we have found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. }

@@

Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach {\color{red} 20} cm during times of high tidal range. {\color{red} As such, storm surge model predictions were corrected by accounting for this error. However, the model will contain tidal energy from shallow water constituents due nonlinear interactions in the model questions. As such, the correction is calculated in the following way:  For each storm surge location of interest, two tidal predictions are calculated: 1) A tidal prediction with all constituents excluding long period constituents (SA, SSA, etc), shallow water constituents, and constituents with signal to noise ratio less than two, and 2) a tidal prediction with only the eight constituents used to force the model.  The difference between these two predictions is added to the model values as a correction. Note that the harmonic analysis to generate the constituents is applied in the same manner as that used in the forcing. That is, the time series from the year of the storm surge is filtered to remove periods of large non-tidal energy and then the harmonic analysis is performed.}  This correction is only applied to the modelled water level and not the modelled residuals. Additionally, since the modelled sea surface height is calculated about a zero reference level, the modelled water level is determined by adding a long term mean sea level at the specified location to the modelled sea surface height. All times are reported in UTC.

@@

In general, there is also very good agreement between the observed and modelled residuals (Figure \ref{fig:feb2006}). The most noticeable difference is {\color{red} a broader peak in the model's surge curve, leading to a delay of a few ours in timing of the peak surge for most of the stations (Table \ref{tab:peak}). Note that the surge manifests as a broad peak in both the modelled and observed residual leading to a high probability that large residuals coincide with the high tide.} As such, it is of minor consequence that the timing of the observed and modelled maximum residual do not match perfectly.

@@

@@

Next, an assessment of the factors that are most important in the storm surge forcing conditions is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}). Several simulations were done: First, a simulation with all the forcing conditions, including tides, local atmospheric forcing (both local winds and local atmospheric pressure), and sea surface height anomaly at the open boundary (also called remote forcing) in the black curve. Second, a simulation with all forcing except the local atmospheric forcing in the dashed green curve. Third, a simulation with all forcing except the remote forcing in the red curve. Fourth, a simulation with only tides and local wind forcing, that is, excluding remote forcing and local atmospheric pressure forcing in the dashed blue curve. {\color{red} Comparisons between these simulations will determine the contributions from the remote forcing, the local wind forcing, and the local atmospheric pressure forcing.} Finally, a simulation without tidal forcing was performed to study the importance of the tide-surge interaction (Figure \ref{fig:tidesurge}). Contributions due to stratification and bathymetric resolution will be left to a future study.

@@

Modelled residuals are generally too low in this hindcast (Figure \ref{fig:dec2006}), however, the timing of the surge agrees well. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The $\gamma^2$ measure suggests that the standard deviation of the error in the modelled residuals is roughly half of the standard deviation of the observed residuals. Note that the performance has degraded when compared to the Feb 2006 and Nov 2006 cases. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point nearby but outside of the domain) is up to 10 cm  lower than the observed residual at Neah Bay, WA (a point nearby but inside of the domain). Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause.

@@

In a simulation without atmospheric forcing, the surge amplitude across the Strait of Georgia is uniform with no pronounced difference between the mainland side and Vancouver Island side (Figure \ref{fig:spatial_sshonly}). While the surge entering the domain from the Pacific Ocean is the leading order contribution, the winds may induce localized changes to the surge amplitude. Interestingly, the winds over the Strait of Juan de Fuca cause a decrease in surge amplitude in this region as the winds are acting to push the water towards the mainland {\color{red}(Figure \ref{fig:spatial})}.  {\color{red} Even without atmospheric forcing, there is still a significant difference in surge amplitude between the Strait of Georgia and the Strait of Juan de Fuca, perhaps the incoming surge wave is partially reflected from the mainland coast in the Strait of Georgia. Note that since local winds are stronger in this case, thelocal wind contribution is more significant than in the Feb 2006 case (6 cm contribution at Point Atkinson). However, this is still a small percentage of the total surge amplitude. }

@@

\subsection{Strong Wind Event}\label{sec:wind}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. Although large wind speeds were observed, the surge associated with this event is rather small when compared to the previous cases (Figure \ref{fig:nov2009}). Two significant surge events are noted, the first occurring on Nov 19 with an observed amplitude of about 50 cm at Point Atkinson, and a second, larger surge of 62.5 cm on Nov 20. {\color{red} The strong wind event of Nov 19 was associated with the smaller of the two surges, supporting the fact that local winds do not contribute significantly to surge amplitudes.}  Neither of the surges occurred during a high tide and so no high water level events were recorded. Although {\color{red} the sea surface heigh plots are} not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy.

>>>>

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modular tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. {\color{red} Although these features are not being used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia.} This work focuses on the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study. {\color{red} NEMO has been successfully used in other coastal modelling studies in the Northeast Atlantic \citep{maraldi2013nemo} and the Mediterranean Sea \citep{brossier2012sensitivity}.}

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait (Figure \ref{fig:domain}). Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells {\color{red} inward from the open boundary. Depths between 0 m and 4 m were set to 4 m and a minimum depth of 4 m was by NEMO.}  As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis. {\color{red} A magnified image of the complex passages through the San Jaun and Gulf Islands is provided in Figure \ref{fig:zoom}.}

@@

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. {\color{red}Tidal predictions were generated with a tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge for 67 constituents. In order to prevent over-predicted constituents due to high surge years, the time series is first filtered by applying a Doodson filter \citep{parker2007NOAA} and removing periods with large non-tidal amplitudes. Long period constituents such as SA, SSA, etc are not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected. } The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

{\color{red} Although Tofino is not the closest tide guage to the western boundary, we have found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. In the future, storm surge forecasts will make use of Neah Bay water level predictions from the National Oceanic and Atmospheric Administration (NOAA).}

@@

Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach {\color{red} 20} cm during times of high tidal range. {\color{red} As such, model water level predictions were corrected by accounting for this error.Note that the model will contain tidal energy from shallow water constituents due nonlinear interactions in the model equations. As such, the correction is calculated in the following way:  For each storm surge location of interest, two tidal predictions are calculated: 1) A tidal prediction with all constituents excluding long period constituents (SA, SSA, etc), shallow water constituents, and constituents with signal to noise ratio less than two, and 2) a tidal prediction with only the eight constituents used to force the model.  The difference between these two predictions is added to the model ses surface height as a correction. Note that the harmonic analysis to generate the constituents is applied in the same manner as that used in the forcing. That is, the time series from the year of the storm surge is filtered to remove periods of large non-tidal energy and then the harmonic analysis is performed. The total modelled water level is determined by adding the mean sea level from the tide gauge record over 1983-2001 to the corrected model sea surface height.} This correction is only applied to the modelled water level and not the modelled residuals. All times are reported in UTC.

@@

In general, there is also very good agreement between the observed and modelled residuals (Figure \ref{fig:feb2006}). {\color{red} The most noticeable difference is a broader peak in the model's surge curve, leading to a delay of a few hours in timing of the peak surge for most of the stations (Table \ref{tab:peak}). Note that the surge manifests as a broad peak in both the modelled and observed residual leading to a high probability that large residuals coincide with the high tide.} As such, it is of minor consequence that the timing of the observed and modelled maximum residuals do not match perfectly.

@@

{\color{red} This case was also used to asses the importance of the northern boundary residual forcing and was used to brielfy examine stratified effects. When the northern boundary forcing is exlcuded the surge amplitude at Campbell River drops by about 10 cm. The other stations were not very much affected. This was also observed in another hindcast (Dec 2006). The effect of stratification was examined by considering a case with an initially horizontally uniform stratification and comparing to the fully stratified case. The largest differences were observed at Victoria, with an average 5 cm increase in sea surface height after an initial adjustment period. Other studies have seen an improvement in modelled surge amplitudes when using a baroclinic mdoel as opposed to a three-dimensional barotropic model \citep{ma2015oceanic}. We plan to investigate stratified effects in more detail in a future study.}

@@

Next, an assessment of the factors that are most important in the storm surge forcing conditions is presented for the Feb 2006 storm surge (Figure \ref{fig:factors}). Several simulations were done: First, a simulation with all the forcing conditions, including tides, local atmospheric forcing (both local winds and local atmospheric pressure), and sea surface height anomaly at the open boundary (also called remote forcing) in the black curve. Second, a simulation with all forcing except the local atmospheric forcing in the dashed green curve. Third, a simulation with all forcing except the remote forcing in the red curve. Fourth, a simulation with only tides and local wind forcing, that is, excluding remote forcing and local atmospheric pressure forcing in the dashed blue curve. {\color{red} Comparisons between these simulations will determine the contributions from the remote forcing, the local wind forcing, and the local atmospheric pressure forcing.} Finally, a simulation without tidal forcing was performed to study the importance of the tide-surge interaction (Figure \ref{fig:tidesurge}).

@@

Modelled residuals are generally too low in this hindcast (Figure \ref{fig:dec2006}), however, the timing of the surge agrees well. All Willmott skill scores rank above 0.88 and the $\gamma^2$ measure also indicates good agreement between model and observations (Table \ref{tab:statistics}). The $\gamma^2$ measure suggests that the standard deviation of the error in the modelled residuals is roughly half of the standard deviation of the observed residuals. Note that the performance has degraded when compared to the Feb 2006 and Nov 2006 cases. This could be due to a number of factors. First, the very large wind speeds observed during this storm are significantly underrepresented by the atmospheric model leading to lower wind stress and less piling of water against the coast. Second, the anomaly forcing at the open boundaries could be too low. The observed residual at Tofino, BC (a point nearby but outside of the domain) is up to 10 cm  lower than the observed residual at Neah Bay, WA (a point nearby but inside of the domain). {\color{red} Since the storm surge is less sensitive to wind forcing at the surface than anomaly forcing at the open boundaries, the latter is likely the cause, although it still does not account for all of the error.}

@@

In a simulation without atmospheric forcing, the surge amplitude across the Strait of Georgia is uniform with no pronounced difference between the mainland side and Vancouver Island side (Figure \ref{fig:spatial_sshonly}). While the surge entering the domain from the Pacific Ocean is the leading order contribution, the winds may induce localized changes to the surge amplitude. Interestingly, the winds over the Strait of Juan de Fuca cause a decrease in surge amplitude in this region as the winds are acting to push the water towards the mainland {\color{red}(Figure \ref{fig:spatial})}.  {\color{red} Even without atmospheric forcing, there is still a significant difference in surge amplitude between the Strait of Georgia and the Strait of Juan de Fuca, perhaps the incoming surge wave is partially reflected from the mainland coast in the Strait of Georgia. Note that since local winds are stronger in this case, the local wind contribution is more significant than in the Feb 2006 case (6 cm contribution at Point Atkinson). However, this is still les than 10 \% of the total surge amplitude. }

@@

\subsection{Strong Wind Event}\label{sec:wind}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. {\color{red} Wind speeds at Point Atkinson reached up to 18 m s$^{-1}$ which is approximately the same magnitude to that observed on Feb 4, 2006.}  Although large wind speeds were observed, the surge associated with this event is rather small when compared to the previous cases (Figure \ref{fig:nov2009}). Two significant surge events are noted, the first occurring on Nov 19 with an observed amplitude of about 50 cm at Point Atkinson, and a second, larger surge of 62.5 cm on Nov 20. {\color{red} The strong wind event of Nov 19 was associated with the smaller of the two surges, supporting the fact that local winds do not contribute significantly to surge amplitudes.}  Neither of the surges occurred during a high tide and so no high water level events were recorded. Although {\color{red} the sea surface height plots are} not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy.
